---
title: "Solutions to Exercise for Chapter 2 of 'What If' by Hérnan and Robins"
author: "Rune H B Christensen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
editor_options:
  chunk_output_type: console
---

## Attach R-packages

```{r}
library(data.table) # for data handling
```

## Questions

### Q1 Type data from table 2.2 into **R**

```{r}
A <- c(0,0,0,0,1,1,1,1,0,0,0,1,1,1,1,1,1,1,1,1)
Y <- c(0,1,0,0,0,0,0,1,1,1,0,1,1,1,1,1,1,0,0,0)
L <- rep(c(0, 1), c(8, 12))
d0 <- data.table(L, A, Y)
d1 <- copy(d0)
d1[]
```

### Q2 Compute the average causal risk for each treatment assuming a conditionally randomized experiment "by hand" using the method of standardization.

Formula for standardization:

$$P[Y^a = 1] = \sum_l P[Y^a = 1  \mid L = l] P[L = l] = \sum_l  P[Y = 1  \mid A = a, L = l] P[L = l]$$
Compute $P[Y = 1 | A = a, L = l]$:
```{r}
d2 <- d1[, .(n = .N, Y. = sum(Y), EY = mean(Y)), .(L, A)]
d2[]
```

Here `n` is the number of rows in the data, `Y.` is the sum of $Y$ and `EY` is the conditional mean (probability of 1) by each level of $A$ and $L$.

Compute $P[L = l]$:
```{r}
d2[, PL := sum(n) / nrow(d1), L]
d2[]
```

Compute the risk for each $a$, $P[Y^a = 1] = \sum_l  P[Y = 1 | A = a, L = l] P[L = l]$:
```{r}
d2[, .(risk = sum(EY * PL)), A]
```

### Q3 Repeat question 2. but using the method of inverse probability weighting instead.

Formula for inverse probability weighting:

$$P[Y^a = 1] = \frac{\sum_i Y_i W_i}{\sum_i W_i}, \quad W_i = 1 / f(A_i \mid L_i) = 1/P[A=a \mid L = l] $$

First step is to compute $f(A_i | L_i)$.
```{r}
d1[, addmargins(table(L, A), margin = 2)]
```

For example, for $L=1$ we have that 

- $f(A = 0 | L = 1) = 3 / 12 = 1/4$
- $f(A = 1 | L = 1) = 9 / 12 = 3/4$

We start by computing the no. cases where $A = 0$ and $A = 1$ for each level of $L$ (the numerator):
```{r}
d1[A == 0, nA := sum(A == 0), L]
d1[A == 1, nA := sum(A == 1), L]
## Or simply:
d1[, nA := .N, .(A, L)]
```

Then compute the number the denominator:
```{r}
d1[, nL := .N, L]
```

Then compute all 4 cases of $f(A_i \mid L_i)$ and finally the weights $W_i = 1 / f(A_i \mid L_i)$:
```{r}
d1[, fAL := nA / nL]
d1[, W := 1 / fAL]
d1
```

Illustrate each of the 4 sections of data:
```{r}
tmp <- d1[, print(cbind(A, L, .SD[])), .(L, A)]
```

Finally we compute the inverse probability weighted risks:
```{r}
d1[, .(risk = sum(Y * W) / sum(W)), A]
```

Alternatively use the inbuilt weighted mean function:
```{r}
d1[, .(weighted_mean = weighted.mean(Y, W)), A]
```


### Q4 Repeat question 2. but generalize the approach in preparation for more complicated situations by using modelling methods such as linear models (R: lm) and predictions

First compute the number of observations within each combination of $A$ and $L$, and $P[L=l]$:
```{r}
d2 <- d0[, .(n = .N), .(L, A)]
d2[, PL := sum(n) / nrow(d1), L]
```

Then use a model to compute $P[Y = 1 \mid A = a, L = l]$:
```{r}
fm <- lm(Y ~ A * L, data=d1)
d2[, pred := predict(fm, d2)]

## Alternative:
lr <- glm(Y ~ A * L, binomial, d1)
d2[, pred_lr := predict(lr, d2, type="response")]
d2[]
```

Compute the risk for each $a$, $P[Y^a = 1] = \sum_l  P[Y = 1 \mid A = a, L = l] P[L = l]$:
```{r}
d2[, .(risk = sum(pred * PL)), A]
```


### Q5 Repeat question 3 but generalize the approach in preparation for more complicated situations by using modelling methods such as logistic regression (R: glm) and predictions or fitted values (R: predict or fitted) to estimate the inverse probability weights.

First we fit the logistic regression model to estimate $P(A | L)$:
```{r}
d1 <- copy(d0)
lr <- glm(A ~ L, data=d1, family=binomial)
d1[, EA := fitted(lr)]
```

Then compute $f(A_i | L_i)$ and the weights:
```{r}
d1[, fAL := fifelse(A == 1, EA, 1 - EA)]
d1[, W := 1/fAL]
d1[]
```

Finally compute the weighted means:
```{r}
d1[, weighted.mean(Y, W), A]
```


## Appendix

### **R** session info:

```{r}
sessionInfo()
```











